Inferring stellar rotation periods using Gaussian processes

### 2017-01-23 documents/ms.tex
new koi section

\newcommand{\columbia}{1}
\newcommand{\oxford}{3j}
\newcommand{\uw}{4}
\newcommand{\princeton}{2}
\newcommand{\naigrain}{333}
\newcommand{\nmcquillan}{100}
\newcommand{\nkois}{1114}
\newcommand{\nkoimcq}{271}
\newcommand{\nkoigood}{222}
\newcommand{\nkoibad}{49}
\newcommand{\kepexample}{1430163}
\newcommand{\kepexampleperiod}{4}
\newcommand{\aigrainexampleperiod}{20.8}
\newcommand{\Kepler}{{\it Kepler}}
\newcommand{\kepler}{\Kepler}
\newcommand{\corot}{{\it CoRoT}}
\newcommand{\Ktwo}{{\it K2}}
\newcommand{\ktwo}{\Ktwo}
\newcommand{\TESS}{{\it TESS}}
\newcommand{\LSST}{{\it LSST}}
\newcommand{\Wfirst}{{\it Wfirst}}
\newcommand{\SDSS}{{\it SDSS}}
\newcommand{\PLATO}{{\it PLATO}}
\newcommand{\Teff}{$T_{\mathrm{eff}}$}
\newcommand{\teff}{$T_{\mathrm{eff}}$}
\newcommand{\FeH}{[Fe/H]}
\newcommand{\feh}{[Fe/H]}
\newcommand{\ie}{{\it i.e.}}
\newcommand{\eg}{{\it e.g.}}
\newcommand{\logg}{log \emph{g}}
\newcommand{\dnu}{$\Delta \nu$}
\newcommand{\numax}{$\nu_{\mathrm{max}}$}
\newcommand{\acfRMS}{9.27}
\newcommand{\pgramRMS}{1.25}
\newcommand{\mcmcRMS}{0.46}

@@

In order to test this rotation period inference method on real data,
we applied it to a set of \nkois\ \Kepler\ Object of Interest (KOI)
host stars, \nkoimcq\ of which had rotation periods previously measured by 
the ACF method by \citet{Mcquillan2013}.  We use the pipeline-corrected flux 
(\texttt{pdcsap_flux} column in the \Kepler\ light curve table), median-normalized
and unit-subtracted.  As with the simulated light curves, we randomly subsample each 
light curve by a factor of 30 and split it into segments of about 300 points
for the purposes of evaluating the likelihood.  We also follow the same MCMC
fitting procedure as with the simulated data.

Figure \ref{fig:kois} compares the periods inferred with the GP method to the
ACF-based periods from \citet{Mcquillan2013} for the \nkoimcq\ overlapping KOIs.
This comparison shows generally very good agreement, 
with the exception of a cluster of stars in the lower right of the figure
where the GP method does not recover the ACF period.  
However, rather than being primarily evidence of weakness of the GP method,
this cluster of nominal failures actually reveals an important advantage
for this method---the possibility of quantitatively motivated quality assessment.

We applied our method to \nmcquillan\ stars with rotation periods previously
measured by \citet{Mcquillan2014}.
A feature of \kepler\ light curves that works to our advantage is that they
are naturally broken up into smaller time units by quarterly (three month)
breaks.
As explained above, splitting the data set into quarters, rather than
modelling the entire time series contiguously speeds up the computation time
as inverting several small matrices is faster than inverting one large one.
The \kepler\ quarter divisions are natural places to split the time series
because the spacecraft rotates by ninety degrees every quarter (three months),
placing each star on a new CCD module.
Pixel response functions and background flux differs from pixel-to-pixel and
module-to-module, so noise properties of \kepler\ light curves change every
quarter.
Additionally, changes in the spacecraft's orientation and position during
quarterly re-pointings temporarily affect the temperature of the CCD,
producing systematic features in the light curves at the start of some
quarters.
We model each quarter separately but the parameters of the GP kernel function
are global, {\it i.e.} we do not use a separate period parameter for each
quarter---there is just one period parameter for an entire light curve.
It would be possible to model the time series with a mixture of some global
parameters and some quarter-specific parameters, for example one might expect
that the amplitude of covariance, $A$ or white noise level, $\sigma$ to vary
on a quarterly basis.
However, since there are seventeen quarters this would lead to thirty-seven
parameters, and in the interest of minimising computation time (adding
parameters leads to longer MCMC burn in and convergence time), we choose to
use global parameters only.

Our results show good agreement with the \citet{Mcquillan2014} measurements,
as shown in \ref{fig:mcquillan}.

@@

At this point it is important to highlight the following caveat.
Our method has been tuned to reproduce the rotation periods of the
\citet{Aigrain2015} simulated light curves.
These light curves were simulated using a very good but still imperfect model.
The input parameters and their distributions were chosen to be as realistic as
possible given current observations.
Unfortunately however, there are limited data on some of these parameters for
stars other than the Sun: for example the distrubution of spot lifetimes or
activity cycle length.
Ensuring that we can recover the rotation periods of these simulated light
curves does not necessarily ensure that we can recover the rotation periods of
real \kepler\ light curves with equal accuracy and precision.
Even using priors on the QP kernel parameters that have been derived from the
\citet{Aigrain2015} sample may be dangerous.
This caveat should be taken into consideration when this method is applied to
real data.

>>>>

\newcommand{\columbia}{1}
\newcommand{\oxford}{3j}
\newcommand{\uw}{4}
\newcommand{\princeton}{2}
\newcommand{\naigrain}{333}
\newcommand{\nmcquillan}{100}
\newcommand{\nkois}{1102}
\newcommand{\nkoimcq}{275}
% \newcommand{\nkoigood}{222}
% \newcommand{\nkoibad}{49}
\newcommand{\kepexample}{1430163}
\newcommand{\kepexampleperiod}{4}
\newcommand{\aigrainexampleperiod}{20.8}
\newcommand{\Kepler}{{\it Kepler}}
\newcommand{\kepler}{\Kepler}
\newcommand{\corot}{{\it CoRoT}}
\newcommand{\Ktwo}{{\it K2}}
\newcommand{\ktwo}{\Ktwo}
\newcommand{\TESS}{{\it TESS}}
\newcommand{\LSST}{{\it LSST}}
\newcommand{\Wfirst}{{\it Wfirst}}
\newcommand{\SDSS}{{\it SDSS}}
\newcommand{\PLATO}{{\it PLATO}}
\newcommand{\Teff}{$T_{\mathrm{eff}}$}
\newcommand{\teff}{$T_{\mathrm{eff}}$}
\newcommand{\FeH}{[Fe/H]}
\newcommand{\feh}{[Fe/H]}
\newcommand{\ie}{{\it i.e.}}
\newcommand{\eg}{{\it e.g.}}
\newcommand{\logg}{log \emph{g}}
\newcommand{\dnu}{$\Delta \nu$}
\newcommand{\numax}{$\nu_{\mathrm{max}}$}
\newcommand{\acfRMS}{9.27}
\newcommand{\pgramRMS}{1.25}
\newcommand{\mcmcRMS}{0.46}

@@

In order to test this rotation period inference method on real data,
we applied it to a set of \nkois\ \Kepler\ Object of Interest (KOI)
host stars, \nkoimcq\ of which had rotation periods previously measured by 
the ACF method by \citet{Mcquillan2013}.  We use the pipeline-corrected flux 
(\texttt{pdcsap_flux} column in the \Kepler\ light curve table), median-normalized
and unit-subtracted.  As with the simulated light curves, we randomly subsample each 
light curve by a factor of 30 and split it into segments of about 300 points
for the purposes of evaluating the likelihood.  We also follow the same MCMC
fitting procedure as with the simulated data, using the ACF-based prior as before.

Initially, we also use the same priors on the hyperparameters for the KOIs 
as for the simulated light curves (Table \ref{tab:priors}).  
However, we found that $\ln l$ and $\ln \Gamma$ tended toward slightly different
values than the simulations.  We also found that the allowed hyperparameter
range that we used for the simulations was too large for the KOI population, 
as maybe $\sim$15\% of the fits tended toward corners in the hyperparameter
space, resulting in poor period measurements.  As a result, after this initial
test, we subsequently adjusted the priors and re-fit all the KOIs.  The final 
priors and parameter ranges that we used are in Table \ref{tab:koipriors}.

\begin{table*}
\begin{center}
\caption{Priors and bounds on the natural logarithms of the GP model parameters, 
        for \Kepler\ light curves}
\begin{tabular}{lcc}
Parameter & Prior & Bounds\\
    \hline
    $\ln A$ & $\mathcal N(-13, 5.7)$ & (-20, 0) \\
    $\ln l$ & $\mathcal N(5.0, 1.2)$ & (2, 8) \\
    $\ln \Gamma$ & $\mathcal N(1.9, 1.4)$ & (0, 3) \\
    $\ln \sigma$ & $\mathcal N(-17, 5)$ & (-20, 0) \\
    $\ln P $ & Uniform / ACF-based & ($\ln 0.5, \ln 100$) \\ 
\end{tabular}
\end{center}
\end{table*}
\label{tab:koipriors}

Figure \ref{fig:mcquillan} compares the periods inferred with the GP method to the
ACF-based periods from \citet{Mcquillan2013} for the \nkoimcq\ overlapping KOIs.
This comparison shows generally very good agreement, with only a few exceptions, 
demonstrating that this method works not only on simulated data, but also on 
real data---with the caveat that for any particular data set, some care is needed 
regarding the setting the priors and ranges for the GP hyperparameters.

@@

% At this point it is important to highlight the following caveat.
% Our method has been tuned to reproduce the rotation periods of the
% \citet{Aigrain2015} simulated light curves.
% These light curves were simulated using a very good but still imperfect model.
% The input parameters and their distributions were chosen to be as realistic as
% possible given current observations.
% Unfortunately however, there are limited data on some of these parameters for
% stars other than the Sun: for example the distrubution of spot lifetimes or
% activity cycle length.
% Ensuring that we can recover the rotation periods of these simulated light
% curves does not necessarily ensure that we can recover the rotation periods of
% real \kepler\ light curves with equal accuracy and precision.
% Even using priors on the QP kernel parameters that have been derived from the
% \citet{Aigrain2015} sample may be dangerous.
% This caveat should be taken into consideration when this method is applied to
% real data.
